Large-scale feature selection with Gaussian mixture models for remote sensing images

### 2016-10-14 paper.tex
last changes ...

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature that most  improves the criterion function $J$  is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been reached. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works as the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion value is higher than the best value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion value. Then SFS is called again. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details of the process.

        This SFFS algorithm eventually evaluates more solutions than the SFS algorithm. The results are expected to be better but the trade-off is an increased computational time which is dependent on the complexity of the dataset.

@@

An efficient implementation of the criterion optimization is proposed.
It is based on  the symmetry property of the covariance  matrix and block
inverse  formula~\cite{IMM2012-03274}. In  the following  it is  shown
that the criterion  can be split into  two parts: one that  needs to be
computed for each  tested variable, and one that needs  to be computed
only once per  selection step. For the  cross-validation part, updates
rules are given to derive sub-models  without the necessity to learn a
GMM models for each fold.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}_c^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)}_c & \mathbf{u}_c      \\
            \mathbf{u}_c^t          & \sigma^{(k)}_c \\
            \end{array}\bigg],
        \end{equation}
        where  $\sigma^{(k)}_c$  is  the variance  of  $\mathbf{x}_c$,
        $\mathbf{u}_c$ is  the $k^{th}$  column of the  matrix without
        the          diagonal           element,          \emph{i.e.},
        $\mathbf{u}_{c}(i) =  \boldsymbol{\Sigma}^{(k)}_{c}(i,k)$ with
        $i  \in [1,k-1]$.   Using block  matrix inverse  formulae, the
        inverse of  the covariance  matrix is  given by  the following
        proposition.
        \begin{prop}[Forward update rule for the inverse of the covariance matrix]
        \label{eq:update-inv}
        \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}_c^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A}_c & \mathbf{v}_c \\
                \mathbf{v}_c^t  & \frac{1}{\alpha_c} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}_c     =     (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}     +
        \frac{1}{\alpha_c} (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1} \mathbf{u}_c
        \mathbf{u}_c^t              (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}$,
        $\mathbf{v}_c           =           -           \frac{1}{\alpha_c}
        (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}       \mathbf{u}_c$       and
        $      \alpha_c      =      \sigma^{(k)}_c      -      \mathbf{u}_c^t
        (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}  \mathbf{u}_c$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)}_c)^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

@@

    \subsection{Numerical issues}
    For each  iteration $k$, after  the selection of  optimal features
    w.r.t  the  selected criterion,  the  inverses  of the  covariance
    matrices and their log-determinant  needs to be computed. However,
    the lack of training samples or the highly correlated features may
    induce  a  badly-conditioned  matrix  with  very  small,  or  even
    negative, eigenvalues.   Such values will degrade  drastically the
    estimation of the  inverse and of the log-determinant,  and so the
    numerical stability.

@@

    \subsection{Implementation}
    \label{sec:otb-module}

@@

    \begin{figure} 
        \centering %Pas touche l'indentation !
        \begin{minted}[fontsize=\tiny,bgcolor=bg]{sh} 
otbcli_TrainGMMSelectionApp -io.il hyper.tif \
                            -io.vd reference.shp \
                            -gmm.varnb 20 -gmm.method forward -gmm.crit jm\
                            -gmm.best 1 -gmm.seed 0\
                            -io.out model.txt

@@

\begin{proof}[Proof of proposition~(\ref{eq:update-quad})]
    \begin{alignat*}{3}
    (&\mathbf{x}^{(k)})^t && (\boldsymbol{\Sigma}^{(k)}_c)^{-1} \mathbf{x}^{(k)} \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
        \left[\begin{array}{cc}
            \mathbf{A}_c   & \mathbf{v}_c \\
            \mathbf{v}_c^t & \frac{1}{\alpha_c}
        \end{array}\right]
        \left[\begin{array}{c} \mathbf{x}^{(k-1)} \\ x_k \end{array}\right] \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
            \left[\begin{array}{c} \mathbf{A}_c \mathbf{x}^{(k-1)} + x_k \mathbf{v}_c \\ \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{x_k}{\alpha_c} \end{array}\right] \\
     &= &&(\mathbf{x}^{(k-1)})^t \mathbf{A}_c \mathbf{x}^{(k-1)} + x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} \\
     & &&+ (\mathbf{x}^{(k-1)})^t \mathbf{v}_c x_k + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \\
     & &&+ \frac{1}{\alpha_c} (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{u}_c \mathbf{u}_c^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1}) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}_c^{(k-1)})^{-1} + \alpha_c \mathbf{v}_c \mathbf{v}_c^t) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c ( (\mathbf{x}^{(k-1)})^t \mathbf{v}_c \mathbf{v}_c^t \mathbf{x}^{(k-1)} \\
     & &&+ 2 \frac{x_k}{\alpha_c} \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c^2}) \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c ( (\mathbf{x}^{(k-1)})^t \mathbf{v}_c + \frac{x_k}{\alpha_c} )^2 \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c ( \left[\begin{array}{cc} \mathbf{v}_c^t & \frac{1}{\alpha_c} \end{array}\right] \mathbf{x}^{(k)} )^2
    \end{alignat*}
\end{proof}

>>>>

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature associated to the highest criterion function $J$  is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been reached. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works as the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion value is higher than the best value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion value. Then SFS is called again. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details about the process.

        This SFFS algorithm evaluates more solutions than the SFS algorithm. The results are expected to be better but the trade-off is an increased computational time which is dependent on the complexity of the dataset.

@@

An efficient implementation of  the criterion optimization is detailed
in  the following.   It  is  based on  the  symmetry  property of  the
covariance matrix  and block inverse  formula~\cite{IMM2012-03274}.  It is shown  that the criterion  can be split  into two
parts: one that needs to be computed for each tested variable, and one
that  needs to  be  computed only  once per  selection  step. For  the
cross-validation part,  updates rules  are given to  derive sub-models
without the necessity to learn a GMM models for each fold.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}_c^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)}_c & \mathbf{u}_c      \\
            \mathbf{u}_c^t          & \sigma^{(k)}_c \\
            \end{array}\bigg],
        \end{equation}
        where  $\sigma^{(k)}_c$  is  the variance  of  $x_c$,
        $\mathbf{u}_c$ is  the $k^{th}$  column of the  matrix without
        the          diagonal           element,          \emph{i.e.},
        $\mathbf{u}_{c}(i) =  \boldsymbol{\Sigma}^{(k)}_{c}(i,k)$ with
        $i  \in [1,k-1]$.   Using block  matrix inverse  formulae, the
        inverse of  the covariance  matrix is  given by  the following
        proposition.
        \begin{prop}[Forward update rule for the inverse of the covariance matrix]
        \label{eq:update-inv}
        \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}_c^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A}_c & \mathbf{v}_c \\
                \mathbf{v}_c^t  & \frac{1}{\alpha_c} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}_c     =     (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}     +
        \frac{1}{\alpha_c} (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1} \mathbf{u}_c
        \mathbf{u}_c^t              (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}$,
        $\mathbf{v}_c           =           -           \frac{1}{\alpha_c}
        (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}       \mathbf{u}_c$       and
        $      \alpha_c      =      \sigma^{(k)}_c      -      \mathbf{u}_c^t
        (\boldsymbol{\Sigma}^{(k-1)}_c)^{-1}  \mathbf{u}_c$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)}_c)^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

@@

    \subsection{Numerical issues}

    For each  iteration $k$, after  the selection of  optimal features
    w.r.t  the  selected criterion,  the  inverses  of the  covariance
    matrices and their log-determinant  needs to be computed. However,
    the lack of training samples or the highly correlated features may
    induce  a  badly-conditioned  matrix  with  very  small,  or  even
    negative, eigenvalues.   Such values will degrade  drastically the
    estimation of the  inverse and of the log-determinant,  and so the
    numerical stability.

@@

    \subsection{Implementation}
    \label{sec:otb-module}
    \begin{figure} 
        \centering %Pas touche l'indentation !
        \begin{minted}[fontsize=\tiny,bgcolor=bg]{sh} 
otbcli_TrainGMMSelectionApp -io.il hyper.tif \
                            -io.vd reference.shp \
                            -gmm.varnb 20 -gmm.method forward -gmm.crit jm\
                            -gmm.best 1 -gmm.seed 0\
                            -io.out model.txt

@@

@@

\begin{proof}[Proof of proposition~(\ref{eq:update-quad})]
    \begin{alignat*}{3}
    (&\mathbf{x}^{(k)})^t && (\boldsymbol{\Sigma}^{(k)}_c)^{-1} \mathbf{x}^{(k)} \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
        \left[\begin{array}{cc}
            \mathbf{A}_c   & \mathbf{v}_c \\
            \mathbf{v}_c^t & \frac{1}{\alpha_c}
        \end{array}\right]
        \left[\begin{array}{c} \mathbf{x}^{(k-1)} \\ x_k \end{array}\right] \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
            \left[\begin{array}{c} \mathbf{A}_c \mathbf{x}^{(k-1)} + x_k \mathbf{v}_c \\ \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{x_k}{\alpha_c} \end{array}\right] \\
     &= &&(\mathbf{x}^{(k-1)})^t \mathbf{A}_c \mathbf{x}^{(k-1)} + x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} \\
     & &&+ (\mathbf{x}^{(k-1)})^t \mathbf{v}_c x_k + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t \Big((\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \\
     & &&+ \frac{1}{\alpha_c} (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{u}_c \mathbf{u}_c^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1}\Big) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t \Big((\boldsymbol{\Sigma}_c^{(k-1)})^{-1} + \alpha_c \mathbf{v}_c \mathbf{v}_c^t\Big) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c} \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c \Big( (\mathbf{x}^{(k-1)})^t \mathbf{v}_c \mathbf{v}_c^t \mathbf{x}^{(k-1)} \\
     & &&+ 2 \frac{x_k}{\alpha_c} \mathbf{v}_c^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha_c^2}\Big) \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c \Big( (\mathbf{x}^{(k-1)})^t \mathbf{v}_c + \frac{x_k}{\alpha_c} \Big)^2 \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha_c ( \left[\begin{array}{cc} \mathbf{v}_c^t & \frac{1}{\alpha_c} \end{array}\right] \mathbf{x}^{(k)} )^2
    \end{alignat*}
\end{proof}
